Global multivariate model learning from hierarchically correlated data

### 2021-01-26 main.tex
discussion v1

This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,baldassi_fast_2014}. 
In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.

@@

What is this useful for? 
-

>>>>

This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}. 
In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.

@@

In this work, we proposed a method for inferring parameters of an Ornstein-Uhlenbeck process using data that is correlated through an evolutionary tree. 
We kept a very general setting in which data can in principle represent any set of continuous phenotypic traits or potentially discrete sequences if a continuous approximation is made. 
As such, our approach is only methodological, and does not directly investigate any particular application.

We showed that due to the Gaussian and time reversible nature of the OU process, it is possible to write the joint covariance matrix of all data vectors in a simple way. 
The resulting matrix $\mathbb{G}$ consists of block entries that represent covariances between pairs of leaves.
The dependence of these blocks on the coupling matrix $\bm{J}$ characterizing the OU process and on the tree structure can be written explicitly. 
Interestingly, $\mathbb{G}$ only depends on the tree structure through the pairwise branch length $\Delta t_{ij}$ separating leaves.

We then proposed a way to compute the likelihood of the data given the tree and the parameters of the OU process, namely the coupling matrix $\mathbf{J}$ and timescale $\gamma$.
This method relies on computing the eigenvalues and vectors of the joint covariance matrix in an efficient manner. 
Indeed, it is possible to separate this calculation in two steps: the first in which we perform the eigen-decomposition of the matrix $\mathbf{J}$, and the second in which we compute eigenvalues and vectors of matrices $\mathbf{G}^a$ that embed the tree structure. 
This reduces the computational complexity from $\mathcal{O}(L^3N^3)$ for a naive inversion of $\mathbb{G}$ to $\mathcal{O}(L^3) + \mathcal{O}(LN^3)$. 
We also show that this method can be used to compute the gradient of the likelihood with respect to parameters with the same complexity.
This makes the problem of inferring $\mathbf{J}$ amenable to maximum likelihood methods using a gradient ascent approach.

Finally, we showed that this process gives expected results on simulated data, with a more accurate reconstruction of parameters than if empirical estimation was performed. 
These simulations highlight the fact that this method is only useful in the intermediate regime of phylogenetic correlations. 
If $\gamma$ is too high, correlation of data points through the tree is weak and an empirical estimation performs well. 
On the other hand, a very low $\gamma$ results in strong phylogenetic biases that make recovering $\mathbf{J}$ impossible.
However, in the intermediate regime, our re-construction of $\mathbf{J}$ that takes the tree into account results in clear benefits over an empirical estimation.

A limitation of our approach remains the long computational time. 
Even with the efficient computation of the gradient, it was necessary to use small system sizes, $L=10$ at most, to repeat our inference process many times with simulated data in a reasonable time. 
For this reason, the framework proposed here is limited to a small number of variables. 
In this respect, it is interesting to note that a different manner of computing the likelihood developed in \cite{mitov_fast_2020} and based on Gaussian integrations on every branch of the tree results in an asymptotic complexity of $\mathcal{O}(NL^3)$.

Although our method can in principle be used for any set of traits, a major motivation in developing it is its potential application to model of proteins sequences. 
Several results in the last years have shown that selection forces shaping the evolution of protein sequences are well described by a pairwise potential. 
The estimation of this potential is performed using homologous sequences, and is therefore biased by the phylogenetic relations between these sequences.  
Results presented here are a first step in disentangling effects due to phylogeny from effects due to selection in a principled way.

However, there remain several challenges in using this framework for protein sequences. 
First, the computational power required to process actual sequences is much larger than what was needed for the small simulated systems presented here. 
As an example, a protein of length $L=100$ will be represented by $q\times 100=2000$ Gaussian variables, where $q=20$ is the number of amino acids. 
This is of course much larger than the $L=10$ system used as an example to test our approach.

A second question is the capacity of a continuous variable approximation, necessary when using Ornstein-Uhlenbeck dynamics, to represent dynamical properties of the landscape protein sequences evolve in. 
This type of approximation has been successfully used before, but in quite different contexts \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}. 
Its use in the context of modelling the evolutionary dynamics of protein sequences remains an open question.
